---
title: "R Notebook for BCB420 Assigment 1"
output: html_notebook
---
**What are the control and test conditions of the dataset? **

In this experiment the RNAseq analysis was use to understand how USP34 works to stimulate osteogenesis. In this experiment the human mesenchymal stem cells were treated with 2 test conditions. One condition was a treatment of siRNA(small interfering Ribonucleic Acid) as a control to inhibit USP34 expression. The other was  a treatment with USP34(The test condition) the test to see what is and isn't expressed in the cells.


**Why is the dataset of interest to you?**
I took a course in comparative vertebrate anatomy and learned about how bone and teeth formation form in various vertebrate taxa. I was curious as to the molecular mechanisms behind strong structures such as bones and dental tissue in terms of their development and maintenance and what genes are related are related to the ones used in these tissues.

#This step allows us to get the required packages for what we need
```{r}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("edgeR", quietly = TRUE))
    BiocManager::install("edgeR")


library("GEOmetadb")
library("edgeR")
library("biomaRt")
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()

file.info('GEOmetadb.sqlite')

```
Downloading the data 
```{r}
sfiles = getGEOSuppFiles('GSE115789')
```

**Ascessing the data**
Here we will look at the data collected and the information it can give us
```{r}
#Information on the File
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
current_gpl_info$title
current_gpl_info$last_update_date
current_gpl_info$organism
fnames = rownames(sfiles)

#number of genes
dim(usp34_exp)

samples <- data.frame(lapply(colnames(usp34_exp)[3:14],
                             FUN=function(x){unlist(strsplit(x, split = "\\."))[c(1)]}))
colnames(samples) <- colnames(usp34_exp)[3:14]
rownames(samples) <- c("condition")
samples <- data.frame(t(samples))
summarized_gene_counts <- sort(table(usp34_exp$gene_id),decreasing = TRUE)

#Value 1 is for the control and value 2 is for the expression
data2plot <- log2(cpm(usp34_exp_filtered[1:100,8:9]))
boxplot(data2plot, xlab = "Samples", ylab = "log2 usp",
          las = 2, cex = 0.5, cex.lab = 0.5,
          cex.axis = 0.5, main = "USP34 RNASeq Samples")
```

**Were there expression values that were not unique for specific genes? How did you handle these?**
The only thing that was measured was the expression value for each gene for each condiditon so there were no unique expression values for any specific gene.

**Were there expression values that could not be mapped to current HUGO symbols?**
Yes, the values used were test_id, gene_id and gene so they will need to all be modified so that they can be mapped to a HUGO symbol.
**How many outliers were removed?**
No outliers were removed



## References
[1] Guo, Yu‐Chen, et al. “Ubiquitin‐Specific Protease USP 34 Controls Osteogenic Differentiation and Bone Formation by Regulating BMP 2 Signaling.” The EMBO Journal, vol. 37, no. 20, 2018, doi:10.15252/embj.201899398.
[2] Isserlin, Ruth. BCB420 - Computational Systems Biology, 22 Jan. 2020, risserlin.github.io/BCB420-lectures-public/lecture3_expression_analysis.html.
[3] Isserlin, Ruth. BCB420 - Computational Systems Biology, 28 Jan. 2020, risserlin.github.io/BCB420-lectures-public/lecture4_data_exploration_and_normalization.html.

